Making Higher-Order Superposition Work

### 2022-03-15 scratch/paper.tex
new bib and paper

The superposition calculus has only recently been extended
to higher-order logic (more precisely, extensional simple type theory
\cite{henkin-1950-completeness}), resulting in
\emph{\lsup} \cite{bbtvw-21-sup-lam}, which we developed
together with Waldmann, as well as \emph{combinatory superposition}
\cite{br-20-full-sup-w-combs} by Bhayat and Reger. Although these two
calculi do not support an interpreted Boolean type,
they can be extended by ad hoc rules \cite{our-bool-paper} that support
most of the Boolean reasoning necessary in practice.

@@

\ourpara{Zipperposition}
Zipperposition \cite{sc-15-simon-phd,bbtvw-21-sup-lam} is a higher-order
theorem prover that implements both \lsup{} variants, combinatory
superposition, and other superposition-like calculi.
The prover was conceived as a testbed for rapidly
experimenting with extensions of first-order superposition, but over time it
has assimilated many of E's techniques and heuristics and become quite powerful.

@@

To support such streams in the given clause procedure, we extend it to
represent the proof state not only by the active ($A$) and passive ($P$) clause
sets, but also by a priority queue $Q$ containing the inference streams,
similar to the ``to do'' set $T$ present in the abstract Zipperposition loop of 
Waldmann et al.\  \cite[Sect.~4]{wtrb-20-sat-framework}.
Each stream is associated with a weight, and $Q$ is sorted in order of
increasing weight\confrep{}{, a low weight corresponding to a high priority}.
When they introduced \lsup, Bentkamp et al.\
\cite{bbtvw-21-sup-lam} described an older version of this extension. Here
we present a newer version in more detail, including heuristics to postpone
unpromising streams. The pseudocode of the modified procedure is as follows:
\vspace{2.5\jot}
%
\newcommand\ParamMode{\ensuremath{K_{\mathrm{fair}}}}
\newcommand\ParamMaxStreams{\ensuremath{K_{\mathrm{best}}}}
\newcommand\ParamRetry{\ensuremath{K_{\mathrm{retry}}}}
\newcommand{\assign}[2]{\State \ensuremath{\mathit{#1} \gets #2}}
\newcommand{\assignSameLine}[2]{\ensuremath{\mathit{#1} \gets #2}}
\algrenewcommand\algorithmicindent{1em}
\begin{algorithmic}[0]
    \Function{ExtractClause}{$Q$, $\mathit{stream}$}
      \assign{maybe\_clause}{\text{pop and compute the first element of } \mathit{stream} }
      \If{$\mathit{stream}$ is not empty}
        \State add $\mathit{stream}$ to $Q$ with an increased weight
      \EndIf
      \State \Return $\mathit{maybe\_clause}$
    \EndFunction

@@

To support higher-order features
such as function extensionality and quantification over functions,
many refutationally complete calculi employ highly prolific rules.
For example, \lsup{} includes a
\infname{FluidSup} rule \cite{bbtvw-21-sup-lam} that very often applies to two
clauses if one of them contains a term of the form $F \, \overline{s}_n$,
where $n > 0$.
We describe three mechanisms to keep rules like these under control.

@@

\newcommand\ParamPenaltyIncrease{\ensuremath{K_\mathrm{incr}}}
Second, \emph{we penalize the streams of expensive inferences}. The weight of
each stream is given an initial value based on characteristics of the inference
premises such as their derivation depth. For prolific rules such as
\infname{FluidSup}, we increment this value by a parameter \ParamPenaltyIncrease. Weights for
less prolific variants of this rule, such as \infname{DupSup} \cite{bbtvw-21-sup-lam}, are increased by a
fraction of $\ParamPenaltyIncrease$ (e.g., $\lfloor \ParamPenaltyIncrease/3 \rfloor$).

@@

\section{Discussion and Conclusion}
\label{sec:ho-tech:discussion}

%  In this paper, we proposed solutions to issues that arise when implementing a
%  higher-order prover based on \lsup. Some of these also apply
%  to combinatory superposition. We also discussed many choices that can be made
%  during proof search and evaluated some of those choices.

\begin{sloppypar}
Back in 1994, Kohlhase \cite[Sect.~1.3]{mk-94-hores} was optimistic about the
future of higher-order automated reasoning:
%
\begin{quote}
  %The author believes that
  The obstacles to proof search intrinsic
  to higher-order logic may well be compensated by the greater expressive power
  of higher-order logic and by the existence of shorter proofs. Thus
  higher-order automated theorem proving will be practically as feasible as
  first-order theorem proving is now as soon as the technological backlog is made up.
\end{quote}
%
For higher-order superposition, the backlog consisted of designing calculus
extensions, heuristics, and algorithms that mitigate its weaknesses. In
this article, we presented such enhancements, justified their design, and
evaluated them. We explained how each weak point in the
higher-order proving pipeline could be improved, from preprocessing to reasoning
about formulas, to delaying unpromising or explosive inferences, to invoking a
backend. Our evaluation indicates that higher-order superposition is now the
state of the art in higher-order reasoning.
\end{sloppypar}

\looseness=-1
Higher-order extensions of first-order superposition have been
considered %from a theoretical perspective    -- The point is made later in the same para. --JB
by Bentkamp et al.\
\cite{bbtvw-21-sup-lam, bbcw-21-lfho} and Bhayat and Reger
\cite{br-19-restricted-unif, br-20-full-sup-w-combs}. They introduced proof calculi,
proved them refutationally complete, and suggested optional rules, but
they hardly discussed the practical aspects of higher-order superposition. Extensions
of SMT are discussed by Barbosa et al. \cite{brotb-19-ho-smt}.
Bachmair and Ganzinger \cite{bg-1992-nonclausal}, Manna and Waldinger
\cite{mw-1979-nonclausal}, and Murray \cite{nm-1982-nonclausal} have studied
nonclausal resolution calculi.

In contrast, there is a vast literature on practical aspects of first-order
reasoning using superposition and related calculi.
The literature evaluates various procedures and techniques
\cite{hv-09-unifalgs,rsv-15-playing-with-avatar}, literal and term order selection
functions \cite{hrsv-16-selsel}, and clause evaluation functions
\cite{sm-2016-clausesel, gs-20-clausesel}, among others. Our work joins the
select club of papers devoted to practical aspects of higher-order
reasoning
\cite{sb-15-beta,wskb-16-effective-norm,fb-2016-internal-guidance-satallax,benzmueller-et-al-05-can-ho-fo-coop}.

%In contrast, higher-order calculi implemented in competitive provers are usually discussed from a theoretical viewpoint.

%\section{Conclusion and Future Work}
%\label{sec:ho-tech:conclusion}

% We filled in this gap: we describe how
% We described some ways in which the explosion incurred by
% \lsup{} can be kept under control. We showed how native
% support for a Boolean type can dramatically improve a higher-order prover's
% performance. For many choices that can be made during the higher-order proof
% search, we described heuristics and evaluated them. We also designed a
% saturation loop that enumerates infinite sets of higher-order inference
% conclusions, while maintaining the same behavior as the standard given clause
% procedure on first-order clauses.

As a next step, we plan to implement the described techniques in E.
%%% @PETAR: In E 2.6, basically. --JB
% Ehoh \cite{ehoh-section}, the $\lambda$-free higher-order extension of E.
We
expect the resulting prover to be substantially more efficient than
Zipperposition. Moreover, we want to investigate the proofs found by provers
such as CVC4 and Satallax but missed by Zipperposition. Finding the reason
behind why Zipperposition fails to prove specific problems will likely result in useful new techniques.

>>>>

The superposition calculus has only recently been extended
to higher-order logic (more precisely, extensional simple type theory
\cite{henkin-1950-completeness}), resulting in
\emph{\lsup} \cite{bentkamp-et-al-2021-lamsup-journal}, which we developed
together with Waldmann, as well as \emph{combinatory superposition}
\cite{br-20-full-sup-w-combs} by Bhayat and Reger. Although these two
calculi do not support an interpreted Boolean type,
they can be extended by ad hoc rules \cite{our-bool-paper} that support
most of the Boolean reasoning necessary in practice.

@@

\ourpara{Zipperposition}
Zipperposition \cite{sc-15-simon-phd,bentkamp-et-al-2021-lamsup-journal} is a higher-order
theorem prover that implements both \lsup{} variants, combinatory
superposition, and other superposition-like calculi.
The prover was conceived as a testbed for rapidly
experimenting with extensions of first-order superposition, but over time it
has assimilated many of E's techniques and heuristics and become quite powerful.

@@

To support such streams in the given clause procedure, we extend it to
represent the proof state not only by the active ($A$) and passive ($P$) clause
sets, but also by a priority queue $Q$ containing the inference streams,
similar to the ``to do'' set $T$ present in the abstract Zipperposition loop of 
Waldmann et al.\  \cite[Sect.~4]{wtrb-20-sat-framework}.
Each stream is associated with a weight, and $Q$ is sorted in order of
increasing weight\confrep{}{, a low weight corresponding to a high priority}.
When they introduced \lsup, Bentkamp et al.\
\cite{bentkamp-et-al-2021-lamsup-journal} described an older version of this extension. Here
we present a newer version in more detail, including heuristics to postpone
unpromising streams. The pseudocode of the modified procedure is as follows:
\vspace{2.5\jot}
%
\newcommand\ParamMode{\ensuremath{K_{\mathrm{fair}}}}
\newcommand\ParamMaxStreams{\ensuremath{K_{\mathrm{best}}}}
\newcommand\ParamRetry{\ensuremath{K_{\mathrm{retry}}}}
\newcommand{\assign}[2]{\State \ensuremath{\mathit{#1} \gets #2}}
\newcommand{\assignSameLine}[2]{\ensuremath{\mathit{#1} \gets #2}}
\algrenewcommand\algorithmicindent{1em}
\begin{algorithmic}[0]
    \Function{ExtractClause}{$Q$, $\mathit{stream}$}
      \assign{maybe\_clause}{\text{pop and compute the first element of } \mathit{stream} }
      \If{$\mathit{stream}$ is not empty}
        \State add $\mathit{stream}$ to $Q$ with an increased weight
      \EndIf
      \State \Return $\mathit{maybe\_clause}$
    \EndFunction

@@

To support higher-order features
such as function extensionality and quantification over functions,
many refutationally complete calculi employ highly prolific rules.
For example, \lsup{} includes a
\infname{FluidSup} rule \cite{bentkamp-et-al-2021-lamsup-journal} that very often applies to two
clauses if one of them contains a term of the form $F \, \overline{s}_n$,
where $n > 0$.
We describe three mechanisms to keep rules like these under control.

@@

\newcommand\ParamPenaltyIncrease{\ensuremath{K_\mathrm{incr}}}
Second, \emph{we penalize the streams of expensive inferences}. The weight of
each stream is given an initial value based on characteristics of the inference
premises such as their derivation depth. For prolific rules such as
\infname{FluidSup}, we increment this value by a parameter \ParamPenaltyIncrease. Weights for
less prolific variants of this rule, such as \infname{DupSup} \cite{bentkamp-et-al-2021-lamsup-journal}, are increased by a
fraction of $\ParamPenaltyIncrease$ (e.g., $\lfloor \ParamPenaltyIncrease/3 \rfloor$).

@@

% For VSCode
% !TEX root = ./conf.tex

%%% TODO @JASMIN @PETAR: The definition of flat L-resolvent is broken because
%%% of the condition on the variables. --JB
%%% TODO @JASMIN @PETAR: The definition of "tautology" is probably wrong for
%%% BCE. Cf. Kiesl et al.

%%% TODO @PETAR: Check @PETAR below. --JB

%\def\orcid#1{} %{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
%TODO: uncomment before submission:
% \def\orcid#1{\orcidID{#1}}

% \bibliographystyle{splncs04}
% \usepackage{amsthm}
% \usepackage{ntheorem}
\usepackage{flushend}
\usepackage{cite}
\usepackage{etoolbox}
\usepackage{dcolumn}
\usepackage{algorithmicx}
\usepackage{regexpatch}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage[noend]{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage[mathscr]{eucal}
\usepackage{prftree}
\usepackage[utf8]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage[english]{babel}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{tikz}
\usepackage{pgfplots} 
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{subcaption}
%\usepackage{caption}
\usepackage{booktabs}
\usepackage[nomessages]{fp}
\usepackage{enumerate}
%\captionsetup{justification=raggedright,singlelinecheck=false}
\xpatchcmd{\algorithmic}{\labelsep 0.5em}{\labelsep 3em}{\typeout{Success!}}{\typeout{Oh dear!}}

%% BEGIN Times font etc.
\usepackage{mathptmx}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.865]{helvet}
%\usepackage{avant}
\DeclareSymbolFont{letters}{OML}{txmi}{m}{it}
%% END Times font etc.

% Used for displaying a sample figure. If possible, figure files should be
% included in EPS format.

%\usepackage[
%   a4paper,
%   pdftex,
%   pdftitle={SAT-Inspired Eliminations for Superposition},
%   pdfauthor={Petar Vukmirovi\'c, Jasmin Blanchette, and Marijn J.H. Heule},
%   pdfkeywords={},
%   pdfborder={0 0 0},
%   draft=false,
%   colorlinks,linkcolor=blue,citecolor=blue,urlcolor=blue,
%   bookmarksnumbered,
%   bookmarks,
%   bookmarksdepth=2,
%   bookmarksopenlevel=2,
%   bookmarksopen]{hyperref}
% If you use the hyperref package, please uncomment the following line to
% display URLs in blue roman font according to Springer's eBook style:
%\renewcommand\UrlFont{\color{blue}\rmfamily}

\allowdisplaybreaks

% \usepackage[mathlines]{lineno}
% \linenumbersep=30pt
% \linenumbers

% \newcommand*\patchAmsMathEnvironmentForLineno[1]{%
%   \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
%   \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
%   \renewenvironment{#1}%
%%      {\linenomath\csname old#1\endcsname}%
%%      {\csname oldend#1\endcsname\endlinenomath}}% 
% \newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
%   \patchAmsMathEnvironmentForLineno{#1}%
%   \patchAmsMathEnvironmentForLineno{#1*}}%
% \AtBeginDocument{%
% \patchBothAmsMathEnvironmentsForLineno{equation}%
% \patchBothAmsMathEnvironmentsForLineno{align}%
% \patchBothAmsMathEnvironmentsForLineno{flalign}%
% \patchBothAmsMathEnvironmentsForLineno{alignat}%
% \patchBothAmsMathEnvironmentsForLineno{gather}%
% \patchBothAmsMathEnvironmentsForLineno{multline}%
% }

% \usepackage{etoolbox} %% <- for \cspreto, \csappto, \patchcmd, \pretocmd, \apptocmd

% %% Patch 'normal' math environments:
%% \newcommand*\linenomathpatch[1]{%
%%   \cspreto{#1}{\linenomath}%
%%   \cspreto{#1*}{\linenomath}%
%%   \cspreto{end#1}{\endlinenomath}%
%%   \cspreto{end#1*}{\endlinenomath}%
% }
% %% Patch AMS math environments:
%% \newcommand*\linenomathpatchAMS[1]{%
%%   \cspreto{#1}{\linenomathAMS}%
%%   \cspreto{#1*}{\linenomathAMS}%
%%   \csappto{end#1}{\endlinenomath}%
%%   \csappto{end#1*}{\endlinenomath}%
% }

%% %% Definition of \linenomathAMS depends on whether the mathlines option is provided
%% \expandafter\ifx\linenomath\linenomathWithnumbers
%%   \let\linenomathAMS\linenomathWithnumbers
%   %% The following line gets rid of an extra line numbers at the bottom:
%%   \patchcmd\linenomathAMS{\advance\postdisplaypenalty\linenopenalty}{}{}{}
% \else
%%   \let\linenomathAMS\linenomathNonumbers
% \fi

%% \linenomathpatch{equation}
%% \linenomathpatchAMS{gather}
%% \linenomathpatchAMS{multline}
%% \linenomathpatchAMS{align}
%% \linenomathpatchAMS{alignat}
%% \linenomathpatchAMS{flalign}

\title{SAT-Inspired Eliminations for~Superposition}

\author{
\IEEEauthorblockN{Petar Vukmirovi\'c\textsuperscript{1}\orcid{0000-0001-7049-6847},
Jasmin Blanchette\textsuperscript{1,2}\orcid{0000-0002-8367-0936},
Marijn J.H. Heule\textsuperscript{3}\orcid{0000-0002-5587-8801}}\\
\IEEEauthorblockA{%
\textsuperscript{1}Vrije Universiteit Amsterdam, Amsterdam, the Netherlands\\ 
\textsuperscript{2}Universit\'e de Lorraine, CNRS, Inria, LORIA, Nancy, France \\
\textsuperscript{3}Carnegie Mellon University, Pittsburgh, Pennsylvania, United States}
}

% \author{
% \IEEEauthorblockN{Petar Vukmirovi\'c\orcid{0000-0001-7049-6847}}
% \IEEEauthorblockA{Vrije Universiteit Amsterdam}
% \and
% \IEEEauthorblockN{Jasmin Blanchette\orcid{0000-0002-8367-0936}}
% \IEEEauthorblockA{
% Vrije Universiteit Amsterdam, \\
% Max-Planck-Institut f\"ur Informatik, \\
% Universit\'e de Lorraine}
% \and
% \IEEEauthorblockN{Marijn J.H. Heule\orcid{0000-0002-5587-8801}}
% \IEEEauthorblockA{Carnegie Mellon University}
% }

%\hyphenation{pre-process-ing in-process-ing}

% \newcommand\ourparagraph[1]{\noindent{\textbf{\upshape#1.}}}
\newcommand\ourparagraph[1]{\subsection{#1}}

%% TYPESETTING: hacks
\newcommand\medrightarrow{\mathrel{{{\color{black}\relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\rightarrow}}}
\newcommand\medleftarrow{\mathrel{{\color{black}\leftarrow}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex\joinrel{{\color{black}\relbar}}}}
\newcommand\medleftrightarrow{\mathrel{\leftarrow\kern-1.685ex\rightarrow}}
\newcommand\Medrightarrow{\mathrel{{{\color{black}\Relbar}\kern-0.9ex\rlap{\color{white}\ensuremath{\blacksquare}}\kern-0.9ex}\joinrel{\color{black}\Rightarrow}}}
\newcommand\Medleftrightarrow{\mathrel{\Leftarrow\kern-1.685ex\Rightarrow}}

\undef\leftrightarrow %% use \medleftrightarrow instead for iff

\newcommand\scif{\quad\text{if~}}
\newcommand{\HL}{\ensuremath{\mathrm{HL}}}
\newcommand{\PHL}{\ensuremath{\mathrm{HL_\mathrm{p}}}}
\newcommand{\cst}[1]{{\mathsf{#1}}}
\newcommand{\eq}{\mathbin{\approx}}
\newcommand{\noteq}{\mathbin{\not\approx}}
\newcommand{\seq}[1]{\ensuremath{(#1)_{n=1}^{\infty}}}
\newcommand{\impk}[4]{\ensuremath{#1 \impl^{#3}_{#4} #2}}
\newcommand{\infname}[1]{\textsc{#1}}
\newcommand{\namedsimp}[3]{\prftree[d][r]{\relax{\infname{#1}}}{\strut#2}{\strut#3}}
\newcommand{\namedsimpsc}[4]{\prftree[d][r]{\relax{\infname{#1}\scif $#4$}}{\strut#2}{\strut#3}}

\newcommand{\impl}{\ensuremath{\medrightarrow}}
% \newcommand{\bigpimpl}[1]{\ensuremath{\overset{#1}{\rightharpoonup}}}
% \newcommand{\bigfimpl}[1]{\ensuremath{\overset{#1}{\hookrightarrow}}}
\newcommand{\bigpimpl}[1]{\hookrightarrow_\mathrm{p}}
\newcommand{\bigfimpl}[1]{\hookrightarrow}
\newcommand{\eqlit}[2]{\ensuremath{#1 \eq #2}}
\newcommand{\neqlit}[2]{\ensuremath{#1 \noteq #2}}
% \newcommand{\eqneqlit}[2]{\ensuremath{#1 \mathrel{\dot{\eq}} #2}}
\newcommand{\pospredlit}[1]{{#1}}
\newcommand{\negpredlit}[1]{\neglit{#1}}
\newcommand{\neglit}[1]{\neg{\kern.2ex #1}}
\newcommand{\arbpredlit}[1]{(\neg)\kern.2ex\pospredlit{#1}}
\newcommand{\cor}{\ensuremath{\mathrel{\lor}}}
\newcommand{\tuple}[2]{\ensuremath{\vec{#1}_{#2}}}
\newcommand{\tuplen}[1]{\tuple{#1}{n}}
\newcommand{\tupleempty}[1]{\tuple{#1}{}}
\newcommand{\flatres}{\mathbin{\smash{\hbox{\Large$\kern-.1ex\rtimes$}}}}  %% (slightly) asymmetric symbol
\newcommand{\flatresiter}{\mathbin\leadsto}
\newcommand{\withoutpred}[2]{\,\overline{\!#1}_{\!#2}}
\newcommand{\withpred}[2]{#1_{#2}}
\newcommand{\withpredpos}[2]{#1^+_{#2}}
\newcommand{\withpredneg}[2]{#1^-_{#2}}
\newcommand{\binset}[1]{#1{\downarrow}_2}
\newcommand\NumberOK[1]{#1}
\newcommand\NumberNOK[1]{\colorbox{red}{#1}}
\newcolumntype{d}{D{.}{.}{0}}
\newcommand{\confrep}[2]{\begin{conf}#1\end{conf}\begin{rep}#2\end{rep}}

\definecolor{light-gray}{gray}{0.925}
\newcommand\MAX[1]{\smash{\setlength{\fboxsep}{.3ex}\colorbox{light-gray}{\ensuremath{\vphantom{('q}{#1}}}}}

\newtheorem{theorem}{Theorem}{\bfseries}{\slshape}
\newtheorem{example}[theorem]{Example}{\bfseries}{\rmfamily}
\newtheorem{remark}[theorem]{Remark}{\bfseries}{\rmfamily}
\newtheorem{definition}[theorem]{Definition}{\bfseries}{\rmfamily}
\newtheorem{lemma}[theorem]{Lemma}{\bfseries}{\slshape}
\newtheorem{corollary}[theorem]{Corollary}{\bfseries}{\slshape}
% \newtheorem*{proof}{Proof}{\slshape}{\rmfamily}

\newlength{\myskip}
\setlength{\myskip}{0.6ex}
% TYPESETTING: when two environments are next to
% each other, we need to somehow squeeze them
\newcommand{\removemyskip}{\vspace{-\myskip}}

\newenvironment{theoremx}% environment name
{% begin code
  \vspace{\myskip}
  \begin{theorem}
}%
{\end{theorem}
 \vspace{\myskip}}% end code

\newenvironment{examplex}% environment name
{% begin code
  \vspace{\myskip}
  \begin{example}
}%
{\end{example}
 \vspace{\myskip}}% end code

\newenvironment{remarkx}% environment name
{% begin code
  \vspace{\myskip}
  \begin{remark}
}%
{\end{remark}
 \vspace{\myskip}}% end code

\newenvironment{definitionx}% environment name
 {% begin code
   \vspace{\myskip}
   \begin{definition}
 }%
 {\end{definition}
  \vspace{\myskip}}% end code

\newenvironment{lemmax}% environment name
{% begin code
\vspace{\myskip}
\begin{lemma}
}%
{\end{lemma}
\vspace{\myskip}}% end code

\newenvironment{corollaryx}% environment name
{% begin code
\vspace{\myskip}
\begin{corollary}
}%
{\end{corollary}
\vspace{\myskip}}% end code

\newenvironment{proof}% environment name
{% begin code
  \vspace{\myskip}
  \textit{Proof:}%
}%
{\vspace{\myskip}}% end code

% Stolen from Jasmin's paper
\newcommand{\qed}{\hfill{\mbox{\ $\IEEEQED$}}}

\maketitle
\setcounter{footnote}{0}

\newcommand{\ourmodel}{\ensuremath{\mathscr{J}}}

\begin{abstract}
Optimized SAT solvers not only preprocess the clause set, they also transform it
during solving as inprocessing. Some preprocessing techniques have been
generalized to first-order logic with equality. In this \paper, we port
inprocessing techniques to work with superposition, a leading first-order proof
calculus, and we strengthen known preprocessing techniques. Specifically, we
look into elimination of hidden literals, variables (predicates), and blocked
clauses. Our evaluation using the Zipperposition prover confirms that the new
techniques usefully supplement the existing superposition machinery.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

Automated reasoning tools have become much more powerful in the last few
decades thanks to procedures such as conflict-driven clause learning (CDCL)
\cite{MSLM09HBSAT} for propositional logic and superposition
\cite{bg-94-superposition} for first-order logic with equality. However,
the effectiveness of these procedures crucially depends on how the input
problem is represented as a clause set. The clause set
can be optimized beforehand (\emph{preprocessing}) or during the execution of
the procedure (\emph{inprocessing}). In this \paper, we lift several preprocessing and
inprocessing techniques from propositional logic to clausal first-order logic
and demonstrate their usefulness in a superposition prover.

For many years, SAT solvers have used inexpensive clause simplification
techniques such as hidden literal and hidden tautology elimination
\cite{hjb-2010-cl-elim,hjb-2011-big-simplification} and failed literal
detection \cite[Sect.~1.6]{jwf-1995-fld}. We generalize these techniques to
first-order logic with equality
(Sect.~\ref{sec:hidden-literal-based-elimination}). Since the generalization
involves reasoning about infinite sets of literals, we propose restrictions to
make them usable.

\emph{Variable elimination}, based on Davis--Putnam resolution \cite{dp-60-dp}, has
been studied in the context of both propositional logic
\cite{sp-04-niver,cs-00-zres} and quantified Boolean formulas (QBFs)
\cite{ab-2004-re}. The basic idea is to resolve all clauses with negative
occurrences of a propositional variable (i.e., a nullary predicate symbol) against
clauses with positive occurrences and delete the parent clauses. E\'en and
Biere \cite{eb-2005-satpreprocess} refined the technique to identify a subset
of clauses that effectively define a variable and use it to further optimize the
clause set. This latter technique, \emph{variable elimination by substitution},
has been an important preprocessor component in many SAT solvers since its
introduction in 2004.

Specializing second-order quantifier elimination
\cite{go-1992-so-pred-elim,hjo-1996-scan}, Khasidashvili and Korovin
\cite{kk-2016-pe-fol} adapted variable elimination to preprocess first-order
problems, yielding a technique we call \emph{singular predicate elimination}. We
extend their work along two axes (Sect.~\ref{sec:predicate-elimination}): We
generalize E\'en and Biere's refinement to first-order logic, resulting in
\emph{defined predicate elimination}, and explain how both types of predicate
elimination can be used during the proof search as inprocessing.

\begin{qle}
It is well known that clauses containing \emph{pure literals}---predicate
literals whose root symbol occurs only with one polarity in the problem---can
be eliminated. We propose a generalization of this technique, under the name
\emph{quasipure literal elimination}, to allow the elimination of some bipolar
literals as well (Sect.~\ref{sec:quasipure-literal-elimination}).
\end{qle}

The last technique we study is \emph{blocked clause elimination} (Sect.\
\ref{sec:satisfiability-by-clause-elimination}). It is used in both SAT
\cite{jbh-10-BCE} and QBF solvers \cite{bls-11-bloqqer}. Its generalization to first-order
logic has produced good results when used as a preprocessor, especially on
satisfiable problems \cite{ksstb-2017-blockedfol}. We explore more ways to use
blocked clause elimination on satisfiable problems, including using it to
establish equisatisfiability with an empty clause set or as an inprocessing
rule. Unfortunately, we find that its use as inprocessing can compromise the
refutational completeness of superposition.

% Preprocessing has been a crucial part of SAT solvers since 2004, when E\'en and
% Biere introduced bounded variable elimination by substitution~\cite{BVE}.
% They based it on Davis--Putnam resolution \cite{dp-60-dp}, which
% replaces all clauses containing a variable $x$ by all nontautological
% resolvents on $x$, but showed how to exploit definition information
% extracted form the clause set to add only a subset of the nontautological
% resolvents. In this way, many variables can be eliminated without
% growing the clause set. % (in terms of the number of clauses). ---that comes later --JB

% Preprocessing is generally effective, but it can be expensive. As a compromise,
% many solvers use inprocessing:\ interleaving the decision procedure with
% preprocessing steps~\cite{mhb-12-inprocessing}. Top-tier SAT solvers, such as CaDiCaL
% \cite{xxx} and Kissat \cite{xxx} alternate about every second between solving
% and inprocessing. In contrast to preprocessing, inprocessing can also be used
% to simplify the clause set after some important deductions have been made.

% Pre-~and inprocessing are successful beyond propositional logic. For example,
% lifting variable and clause elimination techniques to quantified Boolean
% formulas (QBF) helps solve many formulas that are too difficult in practice for
% decision procedures~\cite{bls-11-bloqqer}. This suggests that richer logics, such as
% first- and higher-order logic, can also benefit from elimination techniques.

% To improve the success rate of superposition provers, we extend several of
% preprocessing and inprocessing techniques from propositional to first-order logic,
% including failed literal detection, hidden literal elimination, hidden
% tautology elimination
% (Sect.~\ref{sec:hidden-literal-based-elimination}). These
% techniques are fairly inexpensive in propositional logic but need to be
% curtailed in first-order logic because they would otherwise involve reasoning
% about an infinite set of literals.

% Khasidashvili and Korovin \cite{kk-2016-pe-fol} generalized Davis--Putnam
% resolution to first-order logic with equality. We call their technique
% \emph{singular predicate elimination}. We also adapt bounded variable
% elimination by substitution, yielding \emph{defined predicate elimination}
% (Sect.~\ref{sec:predicate-elimination}). With some care, we can integrate these
% techniques with superposition without losing refutational completeness.

% Finally, we study a first-order generalization of blocked clause
% elimination~\cite{jbh-10-BCE} (Sect.~\ref{sec:satisfiability-by-clause-elimination}).
% %Unlike the other techniques, it weakens the clause set.
% If all clauses in the current clause set are blocked, this means that it is
% satisfiable.
% %It can be used to show
% %satisfiability by establishing equisatisfiability between the initial clause
% %set and the empty clause set.
% We show how this technique can help superposition provers detect
% satisfiability early instead of diverging. We also show, however, that
% interleaving the technique with superposition compromises refutational
% completeness.

All techniques are implemented in the Zipperposition prover
(Sect.~\ref{sec:implementation}), allowing us to ascertain their usefulness
(Sect.~\ref{sec:evaluation}). The best configuration solves \NumberOK{160}
additional problems on benchmarks consisting of all 13\,495 first-order TPTP theorems
\cite{gs-17-tptp}. The raw experimental data are publicly available.%
\footnote{\url{https://doi.org/10.5281/zenodo.4552499}}\confrep{
More details, including all the proofs, can be found in a
technical report \cite{our-report}.}{}

\begin{rep}
This \paper{} expands on our FMCAD 2021 paper \cite{fmcad}. It includes more
examples, lemmas, and proofs. It also corrects a mistake in the handling of
variables in Definition~\ref{def:equality-blocked}.
\end{rep}

\section{Preliminaries}
\label{sec:preliminaries}

\ourparagraph{Clausal First-Order Logic}
\looseness=-1
Our setting is many-sorted, or many-typed, first-order logic \cite{jg-1987-logic-textbook} with
interpreted equality and a distinguished type (or sort) $o$.
Each variable $x$ is assigned a non-Boolean
type, and each symbol $\cst{f}$ is assigned a tuple $(\tau_1, \ldots, \tau_n,
\tau)$ where $n \geq 0$, $\tau_i$ are non-Boolean types, and $\tau$ is
the \emph{result type}. We distinguish between
\emph{predicate symbols}, with $o$ as the result type, and \emph{function symbols}.
Nullary function symbols are called \emph{constants}. Terms are either
variables~$x$ or \confrep{well-typed }{}applications $\cst{f}(t_1, \ldots,
t_n)$\confrep{, or $\cst{f}$ if $n = 0.$}{, of type $\tau$ where $\cst{f}$ is assigned $(\tau_1, \ldots,
\tau_n, \tau)$, and each $t_i$ is a term of type $\tau_i$.
The parentheses are omitted if $n = 0$.} A term is
\emph{ground} if it contains no variables.
We assume standard definitions and
notations for positions, subterms, and contexts \cite{bn-98-tr-and-all-that}.
We abbreviate a vector $(a_1, \ldots, a_n)$ to $\tuplen{a}$ or $\tuple{a}{}$,
and write $\cst{f}^i(s)$ for the $i$-fold application of an unary symbol $\cst{f}$
(e.g., $\cst{f}^3(x) = \cst{f}(\cst{f}(\cst{f}(x)))$).

An atom is an equation $\eqlit{s}{t}$ corresponding to an unordered pair $\{s,
t\}$. A literal is an equation $\eqlit{s}{t}$ or a disequation $\neqlit{s}{t}$.
For every predicate symbol~$\cst{p}$, \confrep{}{the notation }$\cst{p}(\tuple{s}{})$
abbreviates $\eqlit{\cst{p}(\tuple{s}{})}{\top}$, and
$\neglit{\cst{p}(\tuple{s}{})}$ abbreviates
$\neqlit{\cst{p}(\tuple{s}{})}{\top}$, where $\top$ is a distinguished
constant of type $o$. We distinguish between \emph{predicate literals}
$\arbpredlit{\cst{p}(\tuple{s}{})}$ and \emph{functional literals} $\eqlit{s}{t}$,
where $s$ and $t$ are not of type $o$.
Given a literal $L$, we overload notation and write $\neglit{L}$ to
denote its complement. A clause $C$ is a multiset of literals, written as $L_1
\cor \cdots \cor L_n$ and interpreted disjunctively. Clauses are often defined
as sets of literals, but superposition needs multisets; with multisets,
an instance $C\sigma$ always has the same number of literals as $C$, a most
convenient property. Given a clause set $N$, $\binset{N}$ denotes the subset of
its binary clauses: $\binset{N} = \{ L_1 \cor L_2 \mid L_1 \cor L_2 \in N \}$.

\begin{rep}
A substitution $\sigma$ is a well-typed mapping from
variables to terms such that the set $\{x \mid \sigma(x) \not= x \}$ is finite.
%Assuming $x_i \not= t_i$,
Substitutions are written as $\{x_1 \mapsto t_1,
\ldots,\allowbreak x_n \mapsto t_n\}$ or $\{ \tupleempty{x} \mapsto \tupleempty{t} \}$.%
\confrep{}{ A substitution is a \emph{variable renaming} if it is a
bijection from a set of variables to a set of variables.}

We assume the natural extensions of domain, valuation, interpretation and model
(as defined by Fitting \cite{mf-1996-fol}) from unsorted to many-sorted
logic.
Models we consider are \emph{normal}, i.e., they interpret $\eq$ as an
equality relation. Usual notions of (un)satisfiability and (in)validity are
assumed. We write $\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$
satisfies a clause set $N$, for a variable assignment $\xi$. If $\ourmodel$
is a model of $N$ (i.e., it satisfies it under every variable assignment), 
we simply write $\ourmodel \models N.$  Abusing notation,
we write $M \models N$ to denote that $M$ \emph{entails} $N$, i.e., that every
model of $M$ is a model of $N.$
